---
title: "Cartograms of AB109 Population by County"
author: "Moises Evangelista"
date: "August 31, 2015"
output:
  pdf_document:
    fig_width: 7
    fig_height: 9
    fig_caption: true
    fig_crop: false
fontsize: 11pt
header-includes:
- \usepackage{hyperref, fancyhdr}
- \hypersetup{colorlinks,urlcolor=blue}
- \pagestyle{fancy}
- \fancyhead[CO,CE]{\fontsize{10}{12} \selectfont Cartograms of AB109 Population by County}
---

"A cartogram is a map in which some thematic mapping variable such as travel time, population, or Gross National Product is substituted for land area or distance. The geometry or space of the map is distorted in order to convey the information of this alternate variable. There are two main types of cartograms: area and distance cartograms. Cartograms have a fairly long history, with examples from the mid-1800s." from   <https://en.wikipedia.org/wiki/Cartogram>.

For this project, data from the Chief Probation Officers of California (CPOC) website was used. These data contains counts of supervised AB109 population from all 58 counties in California. The AB109 population is composed of Post Release Community Supervision (PRCS) and Offenders with specified felony crimes that are now punishable by local jails. The latter group is also known as 1170h. The cartograms below were developed to visualize AB109 population for each California County. The CPOC data can be accessed [here](http://www.cpoc.org/assets/Realignment/dashboard_county.swf).

```{r options, cache = FALSE, echo=FALSE, include=FALSE}

rm(list=ls(all=TRUE)) #start with empty workspace

startTime <- Sys.time()

# set global chunk options
knitr::opts_chunk$set(cache=TRUE, echo=FALSE, include=FALSE)

setwd("~/GitHub/AB109Cartograms/") # set the working directory
list.files() # see whats there
```


```{r, echo=FALSE, cache=TRUE}

library(RODBC)

channel <- odbcConnectExcel2007("./probationrealignmentdatacpoc.xlsx")
sqlTables(channel)
allContiesData <- sqlFetch(channel, "countydata")
odbcClose(channel)
str(allContiesData); summary(allContiesData)

# remove rows of labels

labelDesc <- allContiesData[1:19, ]
str(labelDesc)

allCountiesData <- allContiesData[-c(1:19), ]
str(allCountiesData)
head(allCountiesData)

# rename this set

names(allCountiesData) <-  c("County","Month","Year","Region",
                             "A1","A2","A3","A4","A5","A7","A8","A9",
                             "B1","B2","B3","B4",
                             "C1")

str(allCountiesData)
summary(allCountiesData)
# add more desc to the variable names

head(labelDesc,20)[,1:2]

names(allCountiesData) <-  c("County","Month","Year","Region",
                             "A1PRCS offenders released","A2PRCS warrant-before (count)",
                             "A3PRCS closures (6 - 12 months; count)",
                             "A4PRCS closures (1 year; count)",
                             "A5PRCS closures (18 months; count)",
                             "A7Active PRCS offenders (population snapshot)",
                             "A8Active PRCS warrant-after (population snapshot)",
                             "A9PRCS Completions",
                             "B11170h (a) jail only sentences (count)",
                             "B2 1170h (b) split sentences (count)",
                             "B3Active Mandatory Supervision (population snapshot)",
                             "B4Mandatory Supervision Completions(Count)",
                             "C1New felony probation grants (count)")

# remove non numeric or character values from names

names(allCountiesData) <- gsub("[^[:alnum:]]", "", names(allCountiesData))

# subset most recent counts
library(dplyr)

lastCounts <- allCountiesData %>%
  filter(Month == "September", Year == "2014")

# some counties do not have current data
# so filter then separately


OtherCounties1 <- allCountiesData %>%
  filter(grepl("Contra Costa|Glenn|Ventura", County),
Month == "December", Year == "2013")

OtherCounties2 <- allCountiesData %>%
  filter(grepl("Alameda|Marin", County),
Month == "March", Year == "2014")

# combine all the sets

comboLastCountyCounts <- rbind(lastCounts, OtherCounties1, OtherCounties2)

# export the data

write.table(comboLastCountyCounts, file = "./comboLastCountyCounts.txt",  sep = "\t", row.names = FALSE)

# using arcgis or qgis merge the counts with 
# california's counties, then use 

# column for PSP counts is comboLas8 
# column for split count is comboLas13
```

```{r}

# after the psps and split county cartograms are 
# plot a map

library(maptools)

califCounties <- readShapePoly("maps/countiesNoIslandsWithCounts")

cartogramSplists <- readShapePoly("maps/cartogramCountiesNoIslandsSplits")

cartogramPSPs <- readShapePoly("maps/cartogramCountiesNoIslandsPsps")

cartogramPopulation <- readShapePoly("maps/countiesNoIslandsWithPopulationCount")

# see this
# https://github.com/hadley/ggplot2/wiki/plotting-polygon-shapefiles

```

```{r cartogramsCombo, fig.width= 7, fig.height = 9, fig.cap= "Los Angeles County has a lower population of 1170(h) population relative to other counties.", message = FALSE, warning = FALSE, include=TRUE} 

library(ggplot2)
library(cowplot)

MapOfCalifCounties <- ggplot(califCounties, aes(long, lat, group = group)) +
  geom_polygon(aes(), fill = "grey40", colour = "grey90", alpha = .7, size = .05) +
  coord_map() +
  theme_minimal() 

CartogramCa1170HPopByCounty <- ggplot(cartogramSplists, aes(long, lat, group = group)) +
  geom_polygon(aes(), fill = "grey40", colour = "grey90", alpha = .7, size = .05) +
  coord_map() +
  theme_minimal()

CartogramCAlifPopulationByCounty <- ggplot(cartogramPopulation, aes(long, lat, group = group)) +
  geom_polygon(aes(), fill = "grey40", colour = "grey90", alpha = .7, size = .05) +
  coord_map() +
  theme_minimal()

CartogramCaPRCSPopByCounty <- ggplot(cartogramPSPs, aes(long, lat, group = group)) +
  geom_polygon(aes(), fill = "grey40", colour = "grey90", alpha = .7, size = .05) +
  coord_map() +
  theme_minimal() 


plot_grid(MapOfCalifCounties,
          CartogramCa1170HPopByCounty,
          CartogramCAlifPopulationByCounty,
          CartogramCaPRCSPopByCounty,
          labels = c("County border",
                     "1170(h) Population",
                     "Resident population",
                     "PRCS population"),
          label_size = 10,
          ncol = 2,
          rel_widths = .5,
          rel_heights = 1)

```

```{r cartogram of splits, fig.width= 7, fig.height = 9, fig.cap= "Cartogram of 1170h population by county", message = FALSE, warning = FALSE, include=TRUE} 

CartogramCa1170HPopByCounty

```

```{r cartogram of Psps, fig.width= 7, fig.height = 9, fig.cap= "Cartogram of PRCS population by county", message = FALSE, warning = FALSE, include=TRUE} 

CartogramCaPRCSPopByCounty
```


```{r map of counties, fig.width= 7, fig.height = 9, fig.cap= "Map of California counties", message = FALSE, warning = FALSE, include=TRUE} 

MapOfCalifCounties

```

```{r map of counties pop, fig.width= 7, fig.height = 9, fig.cap= "Cartogram of California counties' population 2015 estimate", message = FALSE, warning = FALSE, include=TRUE} 


CartogramCAlifPopulationByCounty
```

```{r cache=FALSE}

#get the time
endTime <- Sys.time()
```

\clearpage


The analysis was completed on `r format(Sys.time(), "%a %b %d %X %Y")
` in `r round(difftime(endTime, startTime , units = c( "secs")),0)` seconds. 
